---
title: Cast a line to subsegments in R
subtitle: User-defined function using sf package
header_img: ./assets/img/blog//20190505_benchmarkfunction-1.png
tags: [R,sf,function]
output: 
  md_document:
    variant: markdown_github
    preserve_yaml: true
---


This post introduces a used-defined function used for casting `sf` objects of `class` `LINESTRING` or `POLYGON` into sub-strings.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      dev = "ragg_png")
knitr::opts_knit$set(global.par = TRUE)
library(knitr)
rm(list = ls())
```

## Required R packages

```{r libs, warning=FALSE, message=FALSE, tidy=TRUE}
library(sf)
library(rnaturalearth)
library(dplyr)
```

## The problem

The `sf`package includes [`st_cast`](https://r-spatial.github.io/sf/reference/st_cast.html), a very powerful function that transforms geometries into other different types of geometries (i.e. `LINESTRING`to `POLYGON`, etc.). 

```{r 20190505_italycast, warning=FALSE, message=FALSE, tidy='styler', dev="png",fig.width=500/90, fig.height=500/90, dpi=90, dev.args=list(bg = NA)}
italy = ne_countries(country = "italy", returnclass = "sf")
italy_pol=italy %>% st_cast("POLYGON")
italy_lin=italy_pol %>% st_cast("LINESTRING")
italy_pt=italy_lin %>% st_cast("POINT")
par(mfrow=c(2,2),mar=c(1,1,1,1),bg=NA)
plot(st_geometry(italy),col=c("red","yellow","blue"), main="MULTIPOLYGON")
plot(st_geometry(italy_pol),col=c("red","yellow","blue"), main="POLYGON")
plot(st_geometry(italy_lin),col=c("red","yellow","blue"), main="LINE")
plot(st_geometry(italy_pt),col=c("red","yellow","blue"), main="POINT")

```

What I missed when using `st_cast` is the possibility to "break" the `LINESTRING` objects into sub-segments:

```{r 20190505_italycastsub,warning=FALSE, tidy='styler', echo=FALSE,dev="png",fig.width=250/90, fig.height=250/90, dpi=90 }

par(mfrow=c(1,1),mar=c(1,1,1,1), bg=NA)
stdh_cast_substring = function(x, to = "MULTILINESTRING") {
  ggg = st_geometry(x)
  
  if (!unique(st_geometry_type(ggg)) %in% c("POLYGON", "LINESTRING")) {
    stop("Input should be  LINESTRING or POLYGON")
  }
  for (k in 1:length(st_geometry(ggg))) {
    sub = ggg[k]
    geom = lapply(1:(length(st_coordinates(sub)[, 1]) - 1),
                  function(i)
                    rbind(
                      as.numeric(st_coordinates(sub)[i, 1:2]),
                      as.numeric(st_coordinates(sub)[i + 1, 1:2])
                    )) %>%
      st_multilinestring() %>%
      st_sfc()
    
    if (k == 1) {
      endgeom = geom
    }
    else {
      endgeom = rbind(endgeom, geom)
    }
  }
  endgeom = endgeom %>% st_sfc(crs = st_crs(x))
  if (class(x)[1] == "sf") {
    endgeom = st_set_geometry(x, endgeom)
  }
  
  if (to == "LINESTRING") {
    endgeom = endgeom %>% st_cast("LINESTRING")
  }
  return(endgeom)
}
italy_sub=stdh_cast_substring(italy_lin,"LINESTRING")
plot(st_geometry(italy_sub),col=c("red","yellow","blue"), main="SUBSEGMENT")

```

## An approach

So one possible solution could be to create `LINESTRING` objects for each consecutive pair of `POINT` objects across the original geometry. Let's check it:

```{r 20190505_testspain, warning=FALSE, message=FALSE, tidy='styler', dev="png",fig.width=500/90, fig.height=250/90, dpi=90 }
par(mfrow=c(1,2),mar=c(1,1,1,1))
test = ne_countries(country = "spain", returnclass = "sf") %>%
  st_cast("POLYGON") %>% st_cast("LINESTRING")
plot(st_geometry(test), col = c("red", "yellow", "blue"),main="LINESTRING")

geom = lapply(1:(length(st_coordinates(test)[, 1]) - 1),
              function(i)
                rbind(
                  as.numeric(st_coordinates(test)[i, 1:2]),
                  as.numeric(st_coordinates(test)[i + 1, 1:2])
                )) %>%
  st_multilinestring() %>%
  st_sfc(crs=st_crs(test)) %>%
  st_cast("LINESTRING")
plot(st_geometry(geom), col = c("red", "yellow", "blue"),main="AFTER FUNCTION")

```

## The function `stdh_cast_substring`

Finally, I wrapped the solution into a function and extended it a little bit:

* When the input is not a `LINESTRING` or a `POLYGON` returns an error and stops.

* The function accepts `sf` with several rows or `sfc` objects with several geometries, and returns the same class of input. In the case of `sf` objects, the input `data.frame` is added.

* By default, the output is a `MULTILINESTRING` geometry. This has the benefit that output has the same number of geometries than the input. This can be modified setting the parameter `to` as `LINESTRING`, that in fact only casts the `MULTILINESTRING` object into `LINESTRING`.

```{r function, warning=FALSE, message=FALSE, tidy='styler'}
stdh_cast_substring = function(x, to = "MULTILINESTRING") {
  ggg = st_geometry(x)
  
  if (!unique(st_geometry_type(ggg)) %in% c("POLYGON", "LINESTRING")) {
    stop("Input should be  LINESTRING or POLYGON")
  }
  for (k in 1:length(st_geometry(ggg))) {
    sub = ggg[k]
    geom = lapply(1:(length(st_coordinates(sub)[, 1]) - 1),
                  function(i)
                    rbind(
                      as.numeric(st_coordinates(sub)[i, 1:2]),
                      as.numeric(st_coordinates(sub)[i + 1, 1:2])
                    )) %>%
      st_multilinestring() %>%
      st_sfc()
    
    if (k == 1) {
      endgeom = geom
    }
    else {
      endgeom = rbind(endgeom, geom)
    }
  }
  endgeom = endgeom %>% st_sfc(crs = st_crs(x))
  if (class(x)[1] == "sf") {
    endgeom = st_set_geometry(x, endgeom)
  }
  
  if (to == "LINESTRING") {
    endgeom = endgeom %>% st_cast("LINESTRING")
  }
  return(endgeom)
}

```
 The function could be improved in terms of performance. Given that it works at a coordinate level, for high-resolution objects it has some degree of delay
 
```{r 20190505_benchmarkfunction, warning=FALSE, message=FALSE, tidy='styler', dev="png",fig.width=400/90, fig.height=500/90, dpi=90 }
test100=ne_countries(continent = "south america",
                     returnclass = "sf") %>% 
  st_cast("POLYGON")

test50=ne_countries(50,continent = "south america",
                    returnclass = "sf") %>%
  st_cast("POLYGON")



init=Sys.time()
t1=stdh_cast_substring(test100, "LINESTRING")
end=Sys.time()
kable(end-init,format = 'markdown')


init=Sys.time()
t2=stdh_cast_substring(test50,"LINESTRING")
end=Sys.time()
kable(end-init,format = 'markdown')


par(mfrow=c(1,1),mar=c(0,0,0,0))
plot(st_geometry(test50),col=NA,bg="#C6ECFF")
plot(st_geometry(ne_countries(50,returnclass = "sf")), col="#F6E1B9",border="#646464",add=T)
plot(st_geometry(test50),col="#FEFEE9",border="#646464",add=T)
plot(st_geometry(t2), col = c("red", "yellow", "blue"),add=T, lwd=0.5)

```
 
It can be seen a difference in terms of performance, noting that `test100` has `r length(st_geometry(test100))` polygons decomposed in `r format(length(st_geometry(t1)), nsmall=1, big.mark=",")` sub-strings while `test50` has `r length(st_geometry(test50))` polygons to `r format(length(st_geometry(t2)), nsmall=1, big.mark=",")` sub-strings. In that sense, the original `st_cast` is much faster, although this solution may work well in most cases.
