---
title: "Beautiful Maps with <strong>R</strong> (I): Fishnets, Honeycombs and Pixels"
subtitle: "Awesome simplified maps with R"
tags: [R,beautiful_maps, maps, sf]
header_img: https://dieghernan.github.io/assets/figs/20190602_imgpost-1.png
output: 
  md_document:
    variant: gfm
    preserve_yaml: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(global.par = TRUE)
rm(list = ls())
```

Sometimes you want to produce maps with special layouts. I specially like maps with modified geometries, e.g. simplifiying the original shapes to squares or dots. When doing a little search over the web I found the fantastic post [Fishnets and Honeycomb: Square vs. Hexagonal Spatial Grids](http://strimas.com/spatial/hexagonal-grids/) by [Matt Strimas-Mackey](http://strimas.com/about/) that was a huge inspiration (by the way, don't miss [his blog](http://strimas.com/#posts), full of very interesting pieces of work). 

The only thing I was not completely comfortable with was that the post used the [old-fashioned](https://geocompr.robinlovelace.net/intro.html#the-history-of-r-spatial) `sp` package instead of my personal fav, the `sf` package, (the post was published in 2016, note that author has also moved to the `sf` package since then). So I decided to explore further options with `sf`.

The approach is similar (using grids over a map and work with that) but using [`st_make_grid`](https://www.rdocumentation.org/packages/sf/versions/0.7-4/topics/st_make_grid). I also expanded it by grouping the grids and also producing dots over the geometries. So basically I produced 5 variations of the map:

| Type          | Replacement | Grouped  |
| ------------- |:-------------:| :-----:|
| **Fishnet** | Square | No
| **Puzzle** | Square | Yes
| **Honeycomb** | Hexagon | No
| **Hexbin** | Hexagon | Yes
| **Pixel** | Dot | No


## Required R packages

```{r libs, warning=FALSE, message=FALSE, tidy='styler'}
library(sf)
library(rnaturalearth)
library(dplyr)
library(RColorBrewer)
```

## Working with square grids

Let's use the `square` option of `st_make_grid` and play a bit with it.

```{r 20190602_preparing,  dev.args=list(bg = NA), message=FALSE, warning=FALSE, dpi=90, paged.print=FALSE, tidy='styler', results='hide'}
GB  <- ne_download(50,
                   type = "map_subunits",
                   returnclass = "sf",
                   destdir = tempdir()) %>%
  subset(CONTINENT == "Europe") %>%
  subset(ADM0_A3 == "GBR")

# Projecting and cleaning
GB = st_transform(GB, 3857) %>% select(NAME_EN, ADM0_A3)
initial = GB
initial$index_target = 1:nrow(initial)
target = st_geometry(initial)

#Create my own color palette
mypal = colorRampPalette(c("#F3F8F8", "#008080"))
```

<i class="fas fa-exclamation-triangle"></i> **Warning:** The `cellsize` should be established in the same unit that the projection (in this case is meters). Pay special attention on this, given that if the parameter is too small (meaning that the grid is too dense) **R** could crash easily.
{: .alert .alert-danger .p-3 .mx-2 .mb-3 }

```{r 20190602_squares, warning=FALSE, message=FALSE, tidy='styler',fig.width=500/90, fig.height=350/90, dpi=90, dev.args=list(bg = NA)}
grid = st_make_grid(target,
                    50 * 1000,
                    # Kms
                    crs = st_crs(initial),
                    what = "polygons",
                    square = TRUE
                    )

#To sf
grid = st_sf(index = 1:length(lengths(grid)), grid) # Add index

#We identify the grids that belongs to a entity by assessing the centroid
cent_grid = st_centroid(grid)
cent_merge = st_join(cent_grid, initial["index_target"], left = F)
grid_new = inner_join(grid, st_drop_geometry(cent_merge))

#Fishnet
Fishgeom = aggregate(grid_new,
                     by = list(grid_new$index_target),
                     FUN = min,
                     do_union = FALSE)

#Lets add the df
Fishnet = left_join(Fishgeom %>% select(index_target),
                    st_drop_geometry(initial)) %>%
  select(-index_target)

# Now lets create the Puzzle
Puzzlegeom = aggregate(st_buffer(grid_new, 0.5),#Avoid slivers
                       by = list(grid_new$index_target),
                       FUN = min,
                       do_union = TRUE )# This changes!!!

Puzzle = left_join(Puzzlegeom %>% select(index_target),
                   st_drop_geometry(initial)) %>%
  select(-index_target)

# Plot
par(mfrow = c(1, 2), mar = c(1, 1, 1, 1), bg=NA)
plot(st_geometry(Fishnet), col = mypal(4), main = "Fishnet")
plot(st_geometry(Puzzle), col = mypal(4), main = "Puzzle")

```

## Going hex

Extremely easy. We just need to change the `square` parameter of `st_make_grid` from `TRUE` to `FALSE`

```{r 20190602_hex, warning=FALSE, message=FALSE, tidy='styler',fig.width=500/90, fig.height=350/90, dpi=90,  dev.args=list(bg = NA)}


grid = st_make_grid(target,
                    50*1000, # Kms
                    crs = st_crs(initial),
                    what = "polygons",
                    square = FALSE # This is the only piece that changes!!!
                    )
# Make sf
grid = st_sf(index = 1:length(lengths(grid)), grid) # Add index

#We identify the grids that belongs to a entity by assessing the centroid 
cent_grid = st_centroid(grid)
cent_merge = st_join(cent_grid, initial["index_target"], left = F)
grid_new = inner_join(grid, st_drop_geometry(cent_merge))

#Honeycomb
Honeygeom = aggregate(
  grid_new,
  by = list(grid_new$index_target),
  FUN = min,
  do_union = FALSE
)

#Lets add the df
Honeycomb=left_join(Honeygeom %>%
                    select(index_target),
                  st_drop_geometry(initial)) %>% 
  select(-index_target)

# Now lets create the Hexbin

Hexbingeom=aggregate(
   st_buffer(grid_new,0.5), #Avoid slivers
  by = list(grid_new$index_target),
  FUN = min,
  do_union = TRUE 
  )

Hexbin=left_join(Hexbingeom %>%
                   select(index_target),
                 st_drop_geometry(initial)) %>% 
  select(-index_target)

#Plot
par(mfrow=c(1,2),mar=c(1,1,1,1), bg=NA)
plot(st_geometry(Honeycomb),col=mypal(4), main="Honeycomb")
plot(st_geometry(Hexbin),col=mypal(4), main="Hexbin")
```


## Pixel it!

Also quite easy, just a couple of tweaks more, always using `st_make_grid`.
```{r 20190602_pix, warning=FALSE, message=FALSE, tidy='styler',fig.width=250/90, fig.height=350/90, dpi=90,  dev.args=list(bg = NA)}
grid = st_make_grid(target,
                    50*1000, # Kms
                    crs = st_crs(initial),
                    what = "centers")

# Make sf
grid = st_sf(index = 1:length(lengths(grid)), grid) # Add index

#We identify the grids that belongs to a entity by assessing the centroid 
cent_grid = st_centroid(grid)
cent_merge = st_join(cent_grid, initial["index_target"], left = F)
grid_new = st_buffer(cent_merge, 50*1000/ 2)
Pixelgeom = aggregate(
  grid_new,
  by = list(grid_new$index_target),
  FUN = min,
  do_union = FALSE
)
#Lets add the df
Pixel=left_join(Pixelgeom %>%
                    select(index_target),
                  st_drop_geometry(initial)) %>% 
  select(-index_target)

#Plot
par(mfrow=c(1,1),mar=c(1,1,1,1), bg=NA)
plot(st_geometry(Pixel),col=mypal(4), main="Pixel")
```


## Wrap up

So finally I wrapped all that in a function (see the code in [my repo](https://github.com/dieghernan/dieghernan.github.io/tree/master/assets/functions)), that I named `stdh_gridpol`:
```{r 20190602_function, warning=FALSE, message=FALSE, tidy='styler',fig.width=500/90, fig.height=3*350/90, dpi=90,  dev.args=list(bg = NA)}
stdh_gridpol = function(sf,
                        to = "fishnet",
                        gridsize = as.integer(
                          min(diff(st_bbox(sf)[c(1, 3)]),
                              diff(st_bbox(sf)[c(2, 4)])) /40),
                        sliver = 0.5) {
  if (!unique(st_geometry_type(sf)) %in% c("POLYGON", "MULTIPOLYGON")) {
    stop("Input should be  MULTIPOLYGON or POLYGON")
  }
  if (!to %in% c("fishnet", "puzzle", "honeycomb", "hexbin", "pixel")) {
    stop("'to' should be 'fishnet','puzzle','honeycomb','hexbin' or 'pixel'")
  }
  
  if (class(sf)[1] == "sf") {
    initial = sf
    initial$index_target = 1:nrow(initial)
  } else {
    initial = st_sf(index_target = 1:length(sf), geom = sf)
  }
  
  target = st_geometry(initial)
  
  if (to %in% c("fishnet", "puzzle")) {
    sq = T
  } else {
    sq = F
  }
  if (to == "pixel") {
    grid = st_make_grid(target,
                        gridsize,
                        crs = st_crs(initial),
                        what = "centers")
  } else {
    grid = st_make_grid(
      target,
      gridsize,
      crs = st_crs(initial),
      what = "polygons",
      square = sq
    )
  }
  grid = st_sf(index = 1:length(lengths(grid)), grid) # Add index
  if (to == "pixel") {
    cent_merge = st_join(grid, initial["index_target"], left = F)
    grid_new = st_buffer(cent_merge, gridsize / 2)
  } else {
    cent_grid = st_centroid(grid)
    cent_merge = st_join(cent_grid, initial["index_target"], left = F)
    grid_new = inner_join(grid, st_drop_geometry(cent_merge))
  }
  if (to %in% c("fishnet", "honeycomb", "pixel")) {
    geom = aggregate(
      grid_new,
      by = list(grid_new$index_target),
      FUN = min,
      do_union = FALSE
    )
  } else {
    geom = aggregate(
      st_buffer(grid_new, sliver),
      by = list(grid_new$index_target),
      FUN = min,
      do_union = TRUE
    )
  }
  if (class(initial)[1] == "sf") {
    fin = left_join(geom %>% select(index_target),
                    st_drop_geometry(initial)) %>%
      select(-index_target)
    fin = st_cast(fin, "MULTIPOLYGON")
    return(fin)
  } else {
    fin = st_cast(geom, "MULTIPOLYGON")
    return(st_geometry(fin))
  }
}
# End of the function-----

fish=stdh_gridpol(GB,to="fishnet",gridsize = 50*1000)
puzz=stdh_gridpol(GB,to="puzzle",gridsize = 50*1000)
hon=stdh_gridpol(GB,to="honeycomb",gridsize = 50*1000)
hex=stdh_gridpol(GB,to="hexbin",gridsize = 50*1000)
pix=stdh_gridpol(GB,to="pixel",gridsize = 50*1000)
```

```{r 20190602_functionex, echo=FALSE,  dev.args=list(bg = NA), fig.height=2*350/90, fig.width=750/90, message=FALSE, warning=FALSE, dpi=90, tidy='styler'}
par(mfrow=c(2,3),mar=c(1,1,1,1), bg=NA)
plot(st_geometry(GB),col=mypal(4), main="Initial")
plot(st_geometry(fish),col=mypal(4), main="Fishnet")
plot(st_geometry(puzz),col=mypal(4), main="Puzzle")
plot(st_geometry(hon),col=mypal(4), main="Honeycomb")
plot(st_geometry(hex),col=mypal(4), main="Hexbin")
plot(st_geometry(pix),col=mypal(4), main="Pixel")

```

And that's it! The function `stdh_gridpol` has some alert mechanisms, as accepting only `POLYGON` or `MULTIPOLYGON`, and the default value of `gridsize` is computed in a way that the shortest dimension would have 40 cells.