---
title: "Beautiful Maps with R (II): Fun with flags"
subtitle: "Put a picture on your map"
tags: [r_bloggers,rstats,rspatial,beautiful_maps, maps, raster,flags,sf,function]
last_modified_at: 2020-12-29
header_img: ./assets/img/blog/20190618_imgpost-1.png
excerpt: Want to use a flag (or any *.png file) as a background of your map? You are in the right post.
output: 
  md_document:
    variant: gfm
    preserve_yaml: true
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, tidy = "styler",
  dev = "ragg_png"
)
knitr::opts_knit$set(global.par = TRUE)

rm(list = ls())
par(mar = c(0, 0, 0, 0))
options(gisco_cache_dir = "~/R/mapslib/GISCO/")
```

**Updated 29 december 2020**: All these pieces of work are already available on `cartography >v.2.4.0` on functions [`getPngLayer`](http://riatelab.github.io/cartography/docs/reference/getPngLayer.html). Just install it via `install.packages("cartography")`. A dedicated blog post with examples on this [link](../202002_cartography1).
{: .alert .alert-info .p-3 .mx-2 .mb-3 }

Want to use a flag (or any `*.png` file) as a background of your map? You are in the right post. I am aware that there are some **R** packages out [there](https://github.com/RobertMyles/flagfillr), but we focus here in the option provided by `cartography::getPngLayer()`, that basically converts your image into a raster (see also this article of [Paul Murrell](https://www.stat.auckland.ac.nz/people/pmur002), **"Raster Images in R Graphics"** [(*The R Journal, Volume 3/1, June 2011*)](https://journal.r-project.org/archive/2011/RJ-2011-008/RJ-2011-008.pdf)). 


## Required R packages

```{r libs, warning=FALSE, message=FALSE}
library(dplyr)
library(sf)
library(cartography)
library(mapSpain)
library(giscoR)
```

## Choosing a good source for our shape

In this post I am going to plot a map of Spain with its [autonomous communities (plus 2 autonomous cities)](https://en.wikipedia.org/wiki/Autonomous_communities_of_Spain), that is the first-level administrative division of the country. [Wikipedia](https://en.wikipedia.org/wiki/Autonomous_communities_of_Spain) shows an initial map identifying also the flag of each territory.

For that, I will use `mapSpain`, that uses information from `giscoR`, whose source is the geodata available in [**Eurostat**](https://ec.europa.eu/eurostat). I would also use `giscoR` to get the world around Spain.


```{r 20190618_preparing,  message=FALSE, warning=FALSE,  fig.height=550/90, fig.width=750/90,dpi=90, paged.print=FALSE, results='hide'}

Spain <- esp_get_ccaa(epsg = 3857, res = 3)

World <- gisco_get_countries(epsg = 3857, res = 3)

bboxcan <- esp_get_can_box(epsg = 3857)

# Plot
par(mar = c(0, 0, 0, 0))
plot(st_geometry(Spain),
  col = NA,
  border = NA,
  bg = "#C6ECFF"
)
plot(st_geometry(World),
  col = "#E0E0E0",
  bg = "#C6ECFF",
  add = T
)
plot(st_geometry(Spain), col = "#FEFEE9", add = T)
layoutLayer(
  title = "",
  frame = FALSE,
  scale = 500,
  sources = gisco_attributions(),
  author = "dieghernan, 2019",
)
plot(bboxcan, add = TRUE)
```



Now we have it! A nice map of Spain with a layout based on the [Wikipedia convention for location maps](https://en.wikipedia.org/wiki/Wikipedia:WikiProject_Maps/Conventions/Location_maps).

## Loading the flag

As a first example, I chose [Asturias](https://en.wikipedia.org/wiki/Asturias) to build my code. So the goal here is to create a `RasterBrick` from the desired `*.png` file, add the necessary geographical information and use the shape of Asturias to crop the flag.

```{r 20190618_Asturias,  message=FALSE, warning=FALSE,  fig.width=1000/90,fig.height=(500/2)/90,dpi=90, paged.print=FALSE, tidy='styler', results='hide'}


# 1.Shape---
shp <- Spain %>% filter(ccaa.shortname.es == "Asturias")

# 2.Get flag---

# Masked
url <- "https://upload.wikimedia.org/wikipedia/commons/thumb/3/3e/Flag_of_Asturias.svg/800px-Flag_of_Asturias.svg.png"

flagnomask <- getPngLayer(shp, url, mask = FALSE)

flagmask <- getPngLayer(shp, url, mask = TRUE)

opar <- par(no.readonly = TRUE)
par(mar = c(1, 1, 1, 1), mfrow = c(1, 2))
pngLayer(flagnomask)
plot(st_geometry(Spain), add = T)

# 4.Mask---
pngLayer(flagmask)
plot(st_geometry(Spain), add = T)
par(opar)
```



<i class="fa fa-star"></i> **Pro tip:** Use high-quality `*.png`, otherwise the plot would look quite poor. Here I show an extreme example.
{: .alert .alert-info .p-3 .mx-2 .mb-3 }

```{r 20190618_svg,  message=FALSE, warning=FALSE, dev='png', fig.width=800/(90),fig.height=550/(90),dpi=90, paged.print=FALSE, tidy='styler', results='hide'}


MURshp <- Spain %>% filter(ccaa.shortname.es == "Murcia")
MURLow <- getPngLayer(
  MURshp,
  "https://upload.wikimedia.org/wikipedia/commons/thumb/a/a5/Flag_of_the_Region_of_Murcia.svg/100px-Flag_of_the_Region_of_Murcia.svg.png"
)
MURHigh <- getPngLayer(
  MURshp,
  "https://upload.wikimedia.org/wikipedia/commons/thumb/a/a5/Flag_of_the_Region_of_Murcia.svg/1200px-Flag_of_the_Region_of_Murcia.svg.png"
)


# Plot and compare
opar <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(1, 1, 1, 1))
plot_sf(MURshp, main = "Low")
pngLayer(MURLow, add = TRUE)

plot_sf(MURshp, main = "High")
pngLayer(MURHigh, add = TRUE)

par(opar)
```


Now, we are ready to have [fun with flags](https://the-big-bang-theory.com/fun_with_flags/). It's time to make the flag map of the autonomous communities of Spain.


```{r 20190618_finalmap, tidy='styler', eval=FALSE}

par(mar = c(0, 0, 0, 0), mfrow = c(1, 1))
plot(Spain %>%
  st_geometry(),
col = NA,
border = NA,
bg = "#C6ECFF"
)
plot(st_geometry(World),
  col = "#E0E0E0",
  add = T
)
plot(st_geometry(bboxcan),
  add = T
)
layoutLayer(
  title = "",
  frame = FALSE,
  sources = "© EuroGeographics for the administrative boundaries",
  author = "dieghernan, 2019",
)
# Andalucia
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/9/9a/Bandera_de_Andalucia.svg/1000px-Bandera_de_Andalucia.svg.png"
shp <- Spain %>% filter(ccaa.shortname.es == "Andalucía")
pngLayer(getPngLayer(shp, flag), add = TRUE)

# ...more flags
# Go to the source code of this post on GitHub for the full code

plot(st_geometry(Spain),
  col = NA,
  lwd = 2,
  add = T
)
```

```{r 20190618_allCCAA, echo=FALSE,  fig.height=550/90, fig.width=750/90, message=FALSE, warning=FALSE, dpi=90, paged.print=FALSE, results='hide', eval=TRUE}

par(mar = c(0, 0, 0, 0), mfrow = c(1, 1))
plot(Spain %>%
  st_geometry(),
col = NA,
border = NA,
bg = "#C6ECFF"
)
plot(st_geometry(World),
  col = "#E0E0E0",
  add = T
)
plot(st_geometry(bboxcan),
  add = T
)
layoutLayer(
  title = "",
  frame = FALSE,
  sources = "© EuroGeographics for the administrative boundaries",
  author = "dieghernan, 2019",
)
# Andalucia
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/9/9a/Bandera_de_Andalucia.svg/1000px-Bandera_de_Andalucia.svg.png"
shp <- Spain %>% filter(ccaa.shortname.es == "Andalucía")
pngLayer(getPngLayer(shp, flag), add = TRUE)


# Aragon
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/1/18/Flag_of_Aragon.svg/1200px-Flag_of_Aragon.svg.png"
shp <- Spain %>% filter(ccaa.shortname.es == "Aragón")
pngLayer(getPngLayer(shp, flag, align = "left"), add = TRUE)

# Asturias
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/3/3e/Flag_of_Asturias.svg/1200px-Flag_of_Asturias.svg.png"
shp <- Spain %>% filter(ccaa.shortname.es == "Asturias")
pngLayer(getPngLayer(shp, flag), add = TRUE)


# Cantabria
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/3/30/Flag_of_Cantabria.svg/1200px-Flag_of_Cantabria.svg.png"
shp <- Spain %>% filter(ccaa.shortname.es == "Cantabria")
pngLayer(getPngLayer(shp, flag), add = TRUE)

# Castilla La Mancha

flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/a/a4/Flag_of_Castile-La_Mancha.svg/1200px-Flag_of_Castile-La_Mancha.svg.png"
shp <- Spain[Spain$iso2.ccaa.code == "ES-CM", ]
pngLayer(getPngLayer(shp, flag), add = TRUE)

# Castilla y León
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/1/13/Flag_of_Castile_and_Le%C3%B3n.svg/1200px-Flag_of_Castile_and_Le%C3%B3n.svg.png"
shp <- Spain %>% filter(iso2.ccaa.code == "ES-CL")
pngLayer(getPngLayer(shp, flag), add = TRUE)

# Cataluña
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/c/ce/Flag_of_Catalonia.svg/1200px-Flag_of_Catalonia.svg.png"
shp <- Spain %>% filter(ccaa.shortname.es == "Cataluña")
pngLayer(getPngLayer(shp, flag), add = TRUE)

# Madrid
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/9/9c/Flag_of_the_Community_of_Madrid.svg/1200px-Flag_of_the_Community_of_Madrid.svg.png"
shp <- Spain %>% filter(ccaa.shortname.es == "Madrid")
pngLayer(getPngLayer(shp, flag), add = TRUE)


# Navarra
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/3/36/Bandera_de_Navarra.svg/1200px-Bandera_de_Navarra.svg.png"
shp <- Spain %>% filter(ccaa.shortname.es == "Navarra")
pngLayer(getPngLayer(shp, flag), add = TRUE)

# Comunidad Valenciana
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/d/df/Flag_of_the_Land_of_Valencia_%28official%29.svg/1200px-Flag_of_the_Land_of_Valencia_%28official%29.svg.png"
shp <- Spain %>% filter(iso2.ccaa.code == "ES-VC")
pngLayer(getPngLayer(shp, flag, align = "left"), add = TRUE)

# Extremadura
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/4/4d/Flag_of_Extremadura_%28with_coat_of_arms%29.svg/1200px-Flag_of_Extremadura_%28with_coat_of_arms%29.svg.png"
shp <- Spain %>% filter(iso2.ccaa.code == "ES-EX")
pngLayer(getPngLayer(shp, flag), add = TRUE)

# Galicia
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/6/64/Flag_of_Galicia.svg/1200px-Flag_of_Galicia.svg.png"
shp <- Spain %>% filter(iso2.ccaa.code == "ES-GA")
pngLayer(getPngLayer(shp, flag), add = TRUE)


# Baleares

flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/7/7b/Flag_of_the_Balearic_Islands.svg/1200px-Flag_of_the_Balearic_Islands.svg.png"
shp <- Spain %>% filter(iso2.ccaa.code == "ES-IB")
pngLayer(getPngLayer(shp, flag), add = TRUE)

# La Rioja

flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/d/db/Flag_of_La_Rioja_%28with_coat_of_arms%29.svg/1200px-Flag_of_La_Rioja_%28with_coat_of_arms%29.svg.png"
shp <- Spain %>% filter(iso2.ccaa.code == "ES-RI")
pngLayer(getPngLayer(shp, flag), add = TRUE)

# Basque Country

flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/2/2d/Flag_of_the_Basque_Country.svg/1200px-Flag_of_the_Basque_Country.svg.png"
shp <- Spain %>% filter(iso2.ccaa.code == "ES-PV")
pngLayer(getPngLayer(shp, flag), add = TRUE)

# Murcia
flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/a/a5/Flag_of_the_Region_of_Murcia.svg/1200px-Flag_of_the_Region_of_Murcia.svg.png"
shp <- Spain %>% filter(iso2.ccaa.code == "ES-MC")
pngLayer(getPngLayer(shp, flag), add = TRUE)

# Canarias

flag <-
  "https://upload.wikimedia.org/wikipedia/commons/thumb/b/b0/Flag_of_the_Canary_Islands.svg/1200px-Flag_of_the_Canary_Islands.svg.png"
shp <- Spain %>% filter(iso2.ccaa.code == "ES-CN")
pngLayer(getPngLayer(shp, flag), add = TRUE)


plot(st_geometry(Spain),
  col = NA,
  lwd = 1,
  add = T
)
```

We are done now. If you have suggestion you can leave a comment. As always, if you enjoyed the post you can share it on your preferred social network.

```{r 20190618_imgpost, fig.height=533/90, fig.width=800/90, dpi=90, include=FALSE, dev='png', eval=TRUE}

nuts1 <- gisco_get_nuts(nuts_level = "1", epsg = 3035, res = 3)


noplot <- c("FRY", "ES7", "PT2", "PT3")
NUTS1_Clean <- nuts1 %>%
  filter(CNTR_CODE %in% eurostat::eu_countries$code) %>%
  filter(!NUTS_ID %in% noplot) %>%
  group_by(CNTR_CODE) %>%
  summarise(drop = n())

url <- "https://upload.wikimedia.org/wikipedia/commons/thumb/b/b7/Flag_of_Europe.svg/1200px-Flag_of_Europe.svg.png"

flag <- getPngLayer(NUTS1_Clean, url, mask = FALSE)
flagcut <- getPngLayer(NUTS1_Clean, url, mask = TRUE)
par(mar = c(0, 0, 0, 0))
pngLayer(flag, alpha = 150)
pngLayer(flagcut, bgalpha = 0, add = T)
```

