---
title: "Beautiful Maps with R (III): Patterns and hatched maps"
subtitle: "A solution for b/w and academic maps."
tags: [rstats,rspatial,beautiful_maps,maps,sf,vignette,function,cartography]
last_modified_at: 2020-02-17
header_img: ./assets/img/blog/20191212_imgpost-1.png
output: 
  md_document:
    variant: gfm
    preserve_yaml: true
---

**Updated 17 february 2020**: All these pieces of work are already available on `cartography >v.2.4.0` on functions [`hatchedLayer`](http://riatelab.github.io/cartography/docs/reference/hatchedLayer.html) and [`legendHatched`](https://riatelab.github.io/cartography/docs/reference/legendHatched.html). Just install it via `install.packages("cartography")`. A dedicated blog post with examples on this [link](../202002_cartography1). 
{: .alert .alert-info .p-3 .mx-2 .mb-3 }

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "ragg_png")
knitr::opts_knit$set(global.par = TRUE)
rm(list = ls())
```

On this post I would show how to produce different filling patterns that could be added over your shapefiles with the `cartography` package.


## Required R packages

```{r libs, warning=FALSE, message=FALSE, tidy='styler'}
library(sf)
library(dplyr)
library(giscoR)
library(cartography)
```


```{r 20191212_dotex, message=FALSE, warning=FALSE, dpi=90, fig.width=200/90, fig.height=350/90,  paged.print=FALSE, tidy='styler', results='hide'}
DE <- gisco_get_countries(country = "Germany", epsg = 3857)
```


Let's see how it works.

```{r 20191212_showfun, warning=FALSE, message=FALSE, tidy='styler', fig.height=500/90,fig.width=500/90, dpi=90 }


par(
  mfrow = c(3, 4),
  mar = c(1, 1, 1, 1),
  cex = 0.5
)
hatchedLayer(DE, "dot")
title("dot")
hatchedLayer(DE, "text", txt = "Y")
title("text")
hatchedLayer(DE, "diamond", density = 0.5)
title("diamond")
hatchedLayer(DE, "grid", lwd = 1.5)
title("grid")
hatchedLayer(DE, "hexagon", col = "blue")
title("hexagon")
hatchedLayer(DE, "horizontal", lty = 5)
title("horizontal")
hatchedLayer(DE, "vertical")
title("vertical")
hatchedLayer(DE, "left2right")
title("left2right")
hatchedLayer(DE, "right2left")
title("right2left")
hatchedLayer(DE, "zigzag")
title("zigzag")
hatchedLayer(DE, "circle")
title("circle")
```

Let's play a little bit more with some of the additional features of the function:

```{r 20191212_playing, warning=FALSE, message=FALSE, tidy='styler', fig.height=350/90,fig.width=500/90, dpi=90 }


par(mar = c(1, 1, 1, 1), mfrow = c(2, 3))
plot(st_geometry(DE))
hatchedLayer(
  DE,
  "dot",
  pch = 10,
  density = 0.5,
  cex = 2,
  col = "darkblue",
  add = T
)
plot(st_geometry(DE))
hatchedLayer(
  DE,
  "dot",
  pch = 21,
  col = "red",
  bg = "green",
  cex = 1.25,
  add = T
)
plot(st_geometry(DE), col = "grey")
hatchedLayer(
  DE,
  "text",
  txt = "DE",
  density = 1.1,
  col = "white",
  add = T
)
plot(st_geometry(DE), col = "blue")
hatchedLayer(
  DE,
  "horizontal",
  lty = 3,
  cellsize = 150 * 1000,
  add = T
)
hatchedLayer(DE, "zigzag", lwd = 2, col = "red")
plot(st_geometry(DE), border = "orange", lwd = 2)
hatchedLayer(DE,
  "left2right",
  density = 2,
  col = "orange",
  add = T
)
```

## Adding legends: the `legendHatched` function

As a complementary function, there is also the `legendHatched`. Main parameters are:

-   `pos`, `title.txt`, `title.cex`, `values.cex`,`categ`, `cex` and `frame`: See `?cartography::legendTypo`.
-   `patterns`: vector of patterns to be created for each element on `categ`.
-   `ptrn.bg`: Background of the legend box for each `categ`.
-   `ptrn.text`: Text to be used for each `categ="text"`, as a single value or a vector.
-   `dot.cex`: `cex` of each `categ="dot"`, as a single value or a vector.
-   `text.cex`: text size of each `categ="text"`, as a single value or a vector.
-   As in the case of the `hatchedLayer`function, different graphical parameters can be passed (`lty`, `lwd`, `pch`, `bg` on points).

Note that is also possible to create solid legends, by setting `col` and `ptrn.bg` to the same color. Parameters would respect the order of the `categ` variable.

```{r 20191212_playinglegend, warning=FALSE, message=FALSE,dpi=90, fig.width=500/90, fig.height=500/90,  tidy='styler'}


par(mar = c(0, 0, 0, 0), mfrow = c(1, 1))
plot(st_geometry(DE)) # Null geometry
legendHatched(
  title.txt = "Example 1",
  categ = c("a", "b"),
  patterns = "dot",
  pch = c(16, 23),
  frame = T
)
legendHatched(
  pos = "left",
  title.txt = "Example 2",
  categ = c("c", "d", "other text"),
  patterns = c("text", "zigzag"),
  ptrn.text = c("s", "pp"),
  ptrn.bg = "grey80",
  col = c("red", "blue")
)

legendHatched(
  pos = "topright",
  title.txt = "Example 3",
  categ = c("e", "f", "solid"),
  patterns = c("circle", "left2right"),
  ptrn.bg = c("orange", "yellow", "green"),
  col = c("white", "white", "green"),
  lty = c(2, 4),
  lwd = c(1, 3)
)


legendHatched(
  pos = "bottomright",
  title.txt = "Example 4",
  values.cex = 1.2,
  categ = c("h", "i", "j", "k"),
  patterns = c("grid", "diamond", "horizontal", "dot"),
  cex = 2,
  pch = 22,
  col = "white",
  ptrn.bg = "black",
  bg = "pink"
)
```

I hope that you find this functions useful. Enjoy and nice mapping!

```{r 20191212_imgpost, echo=FALSE, fig.align='center', fig.height=500/90, fig.width=474/90, warning=FALSE, dpi=90, results='hide', tidy='styler', message=FALSE}
library(cartography)

mtq <- st_read(system.file("gpkg/mtq.gpkg", package = "cartography"))
typoLayer(mtq,
  var = "STATUS", legend.pos = "n",
  legend.values.order = c(
    "Prefecture", "Sub-prefecture",
    "Simple municipality"
  ),
  col = c("grey10", "grey50", "grey80"), border = NA
)
mtq$Patts <- cut(mtq$MED, c(-Inf, 15700, Inf), labels = FALSE)
hatchedLayer(mtq[mtq$Patts == 1, ], "left2right",
  density = 2, col = "white", add = TRUE, pch = 3, cex = 0.6
)
hatchedLayer(mtq[mtq$Patts == 2, ], "left2right",
  density = 4, col = "white", add = TRUE
)
legendHatched(
  pos = "bottomleft",
  cex = 1.5,
  values.cex = 0.8,
  title.txt = "Median Income\n(in thousand of euros)",
  categ = c(
    "11.9 - 15.7", "14.7 - 21.8",
    "Prefecture", "Sub-prefecture",
    "Simple municipality"
  ),
  patterns = c("left2right"), density = c(1, 2),
  col = c(rep("black", 2), "grey10", "grey50", "grey80"),
  ptrn.bg = c(rep("white", 2), "grey10", "grey50", "grey80"),
  pch = 3
)
plot(st_geometry(st_union(mtq)), add = TRUE)
```
