---
title: "We'll miss you, UK"
subtitle: "Brexit and the consequences"
tags:
  - rstats
  - rspatial
  - maps
  - flags
  - giscoR
last_modified_at: 2026-10-07
header_img: ./assets/img/blog/20200602_imgpost-1.webp
output:
  md_document:
    variant: gfm
    preserve_yaml: true
excerpt: "Map the United Kingdom’s departure from the EU using country boundaries and flag imagery."
og_image_width: 810
og_image_height: 540
og_image_type: "image/webp"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "ragg_png")
knitr::opts_knit$set(global.par = TRUE)
rm(list = ls())
```


This is just a short post regarding Brexit. Setting aside economic, political
and social considerations, there is another consequence: there is now one fewer
country in the EU.

We will miss you, UK. We wish you the best.

**Historical example:** The mapping examples use **cartography**, which is now
in maintenance mode. Its maintainers recommend [**mapsf**](https://riatelab.github.io/mapsf/)
for new work. The original code and outputs are preserved and have not been
revalidated against current package versions.
{: .alert .alert-info .p-3 .mx-2 .mb-3 }

```{r 20200602_imgpost, fig.height=540/90, fig.width=810/90, message=FALSE, warning=FALSE, dev='ragg_png', dpi=90}
library(cartography)
library(sf)
library(giscoR)


# EU
eu <- giscoR::gisco_countrycode %>% filter(eu)

# Download maps
NUTS1 <- gisco_get_nuts(epsg = 3035, country = eu$ISO3_CODE, nuts_level = 1)


UK <- gisco_get_countries(country = "GBR", epsg = 3035)

noplot <- c("FRY", "ES7", "PT2", "PT3")
NUTS1_Clean <- NUTS1 %>%
  subset(!NUTS_ID %in% noplot) %>%
  group_by(CNTR_CODE) %>%
  summarise(a = dplyr::n())

# Flag image
url <- "https://upload.wikimedia.org/wikipedia/commons/thumb/b/b7/Flag_of_Europe.svg/800px-Flag_of_Europe.svg.png"

# Mask UK
flagcut <- getPngLayer(
  NUTS1_Clean,
  url
)
# Full extent
flag <- getPngLayer(NUTS1_Clean,
  url,
  mask = FALSE
)


par(mar = c(0, 0, 0, 0))
tilesLayer(flag, alpha = 150)
tilesLayer(flagcut, bgalpha = 0, add = T)
plot(st_geometry(UK),
  col = "white",
  border = NA,
  add = TRUE
)
```

