---
title: "Beautiful Maps with R (III): Fun with flags revisited"
subtitle: "Any picture as a basemap"
tags: [rstats,rspatial,beautiful_maps, maps, terra,flags,sf,rasterpic, giscoR]
header_img: ./assets/img/blog/20190618_imgpost-1.png
excerpt: Want to use a flag (or any *.png file) as a background of your map? You are in the right post.
output: 
  md_document:
    variant: gfm
    preserve_yaml: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  tidy = "styler",
  out.width = "100%"
)
rm(list = ls())
```

On 27 Jan. 2022 my package **rasterpic** was accepted on [CRAN](https://cran.r-project.org/package=rasterpic) (Hooray!!). This package allows to geotag images, using an spatial object (from **sf** or **terra**) as a geographic reference.

I tweeted about that, and it seems to have a good feedback from the [#rspatial](https://twitter.com/hashtag/rspatial) community:

<blockquote class="twitter-tweet">

<p lang="en" dir="ltr">

<a href="https://twitter.com/hashtag/rspatial?src=hash&amp;ref_src=twsrc%5Etfw">#rspatial</a> Do we need a package for using our own pictures as base maps? I don't know, but anyway we have it! {rasterpic} 📦 is now on CRAN, and we can convert our pngs to spatial rasters like this <a href="https://t.co/fpoollARoN">https://t.co/fpoollARoN</a> <a href="https://t.co/l9o7rQwdgX">pic.twitter.com/l9o7rQwdgX</a>

</p>

--- dieghernan ن (@dhernangomez) <a href="https://twitter.com/dhernangomez/status/1486666356544225281?ref_src=twsrc%5Etfw">January 27, 2022</a>

</blockquote>

```{=html}
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
```
I received also an interesting reply to this:

<blockquote class="twitter-tweet">

<p lang="en" dir="ltr">

Ooh I think a map with countries filled with their own flag would be poppin'.

</p>

--- Hefin Ioan Rhys (@HRJ21) <a href="https://twitter.com/HRJ21/status/1486976368936116225?ref_src=twsrc%5Etfw">January 28, 2022</a>

</blockquote>

```{=html}
<script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>
```
That remembers me to a [previous post](https://dieghernan.github.io/202002_cartography1/#png-layer) that I wrote when I added some new functions to the **cartography** package, now replaced by the [**mapsf**](https://cran.r-project.org/package=mapsf) package.

With rasterpic we have now an alternative tool for creating maps using images, and this quick post would show you how to do it.

I would replicate the Africa map presented on my previous plot, but this time I would use newer packages, as [**giscoR**](https://cran.r-project.org/package=giscoR) package, and the development version of [**ggspatial**](https://github.com/paleolimbot/ggspatial/pull/92) (not released yet), that adds support to `SpatRaster` object on **ggplot2**. The flags would be extracted from the GitHub repository <https://github.com/hampusborgos/country-flags>.

```{r africa}

# Development version of ggspatial
# devtools::install_github("paleolimbot/ggspatial")
library(ggspatial)
library(ggplot2)
library(giscoR)
library(dplyr)
library(rasterpic)

# For country names
library(countrycode)

world <- gisco_get_countries(epsg = 3857)
africa <- gisco_get_countries(region = "Africa", epsg = 3857)

# Base map of Africa
plot <- ggplot(world) +
  geom_sf(fill="grey90") +
  # Zoom on Africa
  coord_sf(xlim=c(-2000000, 6000000),
           ylim=c(-4000000, 5000000)) +
  theme_minimal() +
  theme(panel.background = element_rect(fill="lightblue"))

plot

```

Now, let's add the flags with a loop:

```{r flag}

# We paste the ISO2 code to each african country
africa$iso2 <- countrycode(africa$ISO3_CODE, "iso3c", "iso2c")

#Get flags from repo - low quality to speed up the code
flagrepo = "https://raw.githubusercontent.com/hjnilsson/country-flags/master/png250px/"

# Loop and add
iso = "AO"
for (iso in africa$iso2[1:2]){
  # Download pic and plot
  imgurl <- paste0(flagrepo, tolower(iso), ".png")
  tmpfile <- tempfile(fileext = ".png")
  download.file(imgurl, tmpfile, quiet = TRUE, mode="wb")
  
  # Raster
  x <- africa %>% filter(iso2 == iso)
  x_rast <- rasterpic_img(x, tmpfile)
  plot <- plot + layer_spatial(x_rast)
  
  
  
  
}


```
