---
title: "Introducing tidyterra"
subtitle: Easily work and ggplot SpatRasters
excerpt: tidyterra provides tidyverse methods for terra objects and geom
  functions for plotting with ggplot2.
tags:
- r_bloggers
- rstats
- rspatial
- maps
- sf
- ggplot2
- tidyterra
- terra
output:
  html_document:
    df_print: paged
  md_document:
    variant: gfm
    preserve_yaml: yes
header_img: ./assets/img/blog/20220525_easteregg-2.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  tidy = "styler",
  out.width = "100%"
)
rm(list = ls())
```

If you have been playing around with **R** for a while, probably you are
familiarized with the `volcano` dataset:

```{r 20220525_volcano}

data("volcano")
image(volcano, col = terrain.colors(256, rev = TRUE))
```

This represents the topographic information of one of the volcanoes of Auckland
(New Zealand), specifically [Maungawhau / Mount
Eden](https://en.wikipedia.org/wiki/Maungawhau_/_Mount_Eden). But **do you know
that this map is flipped?**

On this post I introduce the [tidyterra
package](https://github.com/dieghernan/tidyterra), recently added to
[CRAN](https://CRAN.R-project.org/package=tidyterra) and I would show you how to
geotag the `volcano` dataset. We would produce also ggplot2 maps using the
functions of `tidyterra`.

```{r}
# Libraries
library(terra)
library(ggplot2)
library(tidyterra)
library(maptiles)
library(sf)
```

## Wait, `volcano` is flipped?

Let's check it out. Thanks to the package `maptiles` we can have a glimpse of
the location of Maungawhau using map tiles (as Google Maps uses). We would use
`tidyterra` for displaying the map tile:

```{r 20220525_tile}

# location of Maungawhau

box <- c(
  174.7611552780,
  -36.8799200525,
  174.7682380109,
  -36.8719519780
)
class(box) <- "bbox"
box <- st_as_sfc(box)
st_crs(box) <- 4326

box <- box %>%
  # To crs for NZGD49
  st_transform(27200)

tile <- get_tiles(box, crop = TRUE, zoom = 16)


ggtile <- ggplot() +
  geom_spatraster_rgb(data = tile)

ggtile
```

So well, here you go. A neat and crisp RGB tile of Maungawhau. Now, the next
question is, how to match the `volcano` dataset (a matrix) with this tile (a
geo-tagged map tile)? Let's check it out

## Working with SpatRasters

Thanks to the `terra` package we can start converting `volcano` into a
SpatRaster:

```{r 20220525_volcano_raster}

volcano_rast <- rast(volcano)

terra::plot(volcano_rast)

# Wait, it is flipped!
volcano_rast_ok <- rast(volcano[seq(nrow(volcano), 1, -1), 
                                seq(ncol(volcano), 1, -1) ])

# Much better!
terra::plot(volcano_rast_ok)

volcano_rast_ok
```

Nice! Now we have a raster of `volcano`, but still without geotagged
information. Thanks to this article of Tomislav Hengl
([\@tom_hengl](https://twitter.com/tom_hengl)) we can check the basic geographic
parameters of `volcano` (see [Volcano
Maungawhau](https://geomorphometry.org/volcano-maungawhau/)), that are:

-   **CRS**: EPSG:27200
-   **xllcorner**: 2667400
-   **yllcorner**: 6478700
-   **cellsize**: 10 m
-   **ncols**: 61
-   **nrows**: 87

And we can translate that easily to an empty SpatRaster:

```{r}

# Extra length for proper handling extent
xrange <- range(seq(from = 2667400, length.out = 62, by = 10))
yrange <- range(seq(from = 6478700, length.out = 88, by = 10))

template <- rast(
  crs = "EPSG:27200",
  xmin = xrange[1],
  xmax = xrange[2],
  ymin = yrange[1],
  ymax = yrange[2],
  resolution = 10
)
template
```

So now we only need to transfer the values from `volcano_rast_ok` to our
template:

```{r 20220525_create_volcano2}

# Use tidyterra for pull the values of one raster
# and create a new layer

volcano2 <- template %>%
  mutate(elevation = pull(volcano_rast_ok, lyr.1)) %>%
  select(elevation)

volcano2

terra::plot(volcano2)

# And plot it
ggtile +
  geom_spatraster(data = volcano2) +
  scale_fill_terrain_c(alpha = 0.75)
```

## An Easter egg

The `volcano` dataset may not be completely up to date. As a compliment,
`tidyterra` includes a `.tif` file with the same dimensions that our `volcano2`
raster, but with the topographic values extracted from [Auckland LiDAR 1m DEM
(2013)](https://data.linz.govt.nz/layer/53405-auckland-lidar-1m-dem-2013/) and
resampled to a resolution of 5x5 meters, for package size optimization. See here
how to load it and check the plotting `tidyterra` possibilities:

```{r 20220525_easteregg}

# Load out Easter Egg

volcano2_easter <- rast(system.file("extdata/volcano2.tif",
                                    package = "tidyterra"))

volcano2_easter
terra::plot(volcano2_easter)


# Only altitudes of more than 130m

volcano_filter <- volcano2_easter %>% 
  filter(elevation > 130)


ggtile +
  geom_spatraster(data = volcano_filter) +
  scale_fill_viridis_c(na.value = NA, alpha = 0.7) +
  labs(fill = "Elevation (m)")


# Contour lines

ggtile +
  geom_spatraster_contour(data = volcano2_easter, binwidth = 10)


# Contour lines + contour polygons

ggtile +
  geom_spatraster_contour_filled(
    data = volcano2_easter,
    breaks = seq(70, 210, 20),
    alpha = 0.7
  ) +
  geom_spatraster_contour(
    data = volcano2_easter, binwidth = 2.5,
    alpha = 0.7, size = .2, color = "grey10"
  ) +
  coord_sf(expand = FALSE)
```
