---
title: "Mapping Antarctica"
subtitle: "Cool maps from the South Pole"
tags:
- r_bloggers
- beautiful_maps
- rstats
- rspatial
- maps
- ggplot2
- sf
- giscoR
output:
  html_document:
  md_document:
    variant: gfm
    preserve_yaml: true
header_img: xxx.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  dev = "ragg_png",
  dpi = 300,
  tidy = "styler",
  out.width = "100%",
  fig.cap = ""
)
rm(list = ls())
```

Creating maps with **R** is usually straightforward, but representations that
cross the [International Date
Line](https://en.wikipedia.org/wiki/International_Date_Line) or that use polar
projections can be tricky.

Different spatial-data providers use different conventions: some break
geometries at certain longitudes (for example, cutting the Chukchi Peninsula),
while others omit portions of the data. These inconsistencies can produce
awkward artifacts near the poles.

In this post I fix the [GISCO (European
Commission)](https://ec.europa.eu/eurostat/web/gisco) shapefile for Antarctica
and produce clean orthographic maps. I walk through the manual corrections and
then create a few example maps.

```{r}
# Libraries
library(tidyverse)
library(sf)
library(giscoR)
library(ggrepel)
library(rmapshaper)
```

## Fixing the geometry

First, we obtain the GISCO Antarctica polygon and transform it to an
orthographic projection centered on the South Pole.

```{r 202510_init}
antarct <- gisco_get_countries(year = 2024, resolution = 1, country = "ATA") %>%
  select(NAME = NAME_ENGL) |>
  # Ortho proj centered in the South Pole
  st_transform(crs = "+proj=ortho +lat_0=-90 +lon_0=0")

ggplot(antarct) +
  geom_sf(fill = "lightblue")
```

The shapefile contains a visible "lollipop" cut that looks unnatural in an
orthographic projection. I correct it manually by:

1.  Identify the polygon that represents the main Antarctic landmass.
2.  Convert that polygon to a sequence of coordinates (points).
3.  Remove the small sequence of points that create the artifact.
4.  Rebuild the polygon from the cleaned coordinates and replace the broken
    geometry with the corrected one.

We convert polygons to point coordinates and inspect them to find the offending
sequence:

```{r 202510_init_ant}
# Identify the max
ant_explode <- antarct |>
  st_cast("POLYGON")

nrow(ant_explode)

# Max polygon

ant_max <- ant_explode[which.max(st_area(ant_explode)), ]

coords <- st_coordinates(ant_max) |>
  as_tibble() |>
  # Add id for points
  mutate(np = row_number())


ggplot(coords, aes(X, Y)) +
  geom_point(size = 0.05, color = "darkblue") +
  geom_text(aes(label = np), check_overlap = TRUE) +
  coord_equal()
```

From the plotted indices, we can see the problematic points fall roughly in the
range 8200–9200. We inspect that interval in detail to select the exact indices
to remove.

```{r 202510_test_points}
test <- coords |>
  filter(np %in% seq(8200, 9200))

test |>
  ggplot(aes(X, Y)) +
  geom_point(size = 0.05, color = "darkblue") +
  geom_text(aes(label = np), check_overlap = TRUE)
```

::: {.alert .alert-warning .p-3 .mx-2 .mb-3}
Note: This cleaning is tailored to this specific shapefile and may need to be
repeated for other shapefiles. The approach is straightforward but depends on
the particular geometry and projection.
:::

```{r 202510_final_sol, out.width="50%", fig.show="hold"}
# Final solution after some iterations...

test |>
  filter(np %in% seq(8289, 9130)) |>
  ggplot(aes(X, Y)) +
  geom_point(color = "darkblue") +
  labs(title = "To remove")


test |>
  filter(!np %in% seq(8289, 9130)) |>
  ggplot(aes(X, Y)) +
  geom_point(color = "darkblue") +
  labs(title = "To keep")
```

After removing the offending points, we rebuild the polygon and reconstitute the
full Antarctica shape from the corrected piece plus the remaining polygons.

```{r 202510_good_pol}
# From coordinates to polygon
newpol <- coords |>
  as.data.frame() |>
  filter(!np %in% seq(8289, 9130)) |> # Removing offending points
  select(X, Y) |>
  as.matrix() |>
  list() |>
  st_polygon() |>
  st_sfc() |>
  st_set_crs(st_crs(ant_max))

ant_max_fixed <- st_sf(st_drop_geometry(ant_max), geometry = newpol)

# Regenerate initial shape
antarctica_fixed <- bind_rows(
  ant_max_fixed,
  ant_explode[-which.max(st_area(ant_explode)), ]
) |>
  group_by(NAME) |>
  summarise(m = 1) |>
  select(-m) |>
  st_make_valid()

antarctica_fixed

ggplot(antarctica_fixed) +
  geom_sf(fill = "lightblue")
```

## Plotting examples

With the corrected shape we can produce maps. Below are a few examples based on
proposed Antarctic flag designs.

### Graham Bartram's proposal (1996)

A simple rendition of Bartram's original concept:

```{r 202510_bartram}
bbox <- st_bbox(antarctica_fixed) # For limits on the panel

antarctica_fixed |>
  ggplot() +
  geom_sf(fill = "white", color = NA) +
  theme(
    panel.background = element_rect(fill = "#009fdc"),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(title = "Graham Bartram's proposal") +
  coord_sf(
    xlim = c(bbox[c(1, 3)]) * 1.8,
    ylim = c(bbox[c(2, 4)]) * 1.4
  )
```

### Emblem of the Antarctic Treaty

This example uses graticules to create a concentric "bullseye" pattern around
Antarctica. Generating such graticules and merging meridians requires a few
extra steps to avoid small gaps near the pole.

```{r 202510_treaty1}
# Need graticules
grats <- giscoR::gisco_get_countries() |>
  st_transform(st_crs(antarctica_fixed)) |>
  # Specify the cuts of the graticules
  st_graticule(
    lat = c(-80, -70, -60),
    lon = seq(-180, 180, 30),
    ndiscr = 10000,
    margin = 0.000001
  )


ggplot(grats) +
  geom_sf(color = "darkblue")
```

We merge meridians so the area around the South Pole is filled. `st_graticule()`
can leave a tiny hole at the pole; we fix this by joining complementary
meridians.

```{r 202510_fixgrats}
# Merge meridians
merid <- lapply(seq(-180, 0, 30), function(x) {
  df <- grats |>
    filter(type == "E") |>
    filter(degree %in% c(x, x + 180))

  df2 <- df |>
    st_geometry() |>
    st_cast("MULTIPOINT") |>
    st_union() |>
    st_cast("LINESTRING")

  sf_x <- st_sf(
    degree = x,
    type = "E",
    geometry = df2
  )
}) |> bind_rows()


grats_end <- merid |>
  bind_rows(grats |>
    filter(type != "E"))
```

We then cut and color the resulting graticules so they form the emblem-like
pattern.

```{r 202510_treaty2}
# Cut since some grats should be colored differently

antarctica_simp <- rmapshaper::ms_simplify(antarctica_fixed, keep = 0.005)
grats_yes <- st_intersection(grats_end, antarctica_simp)
grats_no <- st_difference(grats_end, antarctica_simp)

antarctica_simp |>
  ggplot() +
  geom_sf(fill = "white", color = NA) +
  theme(
    panel.background = element_rect(fill = "#072b5f"),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  geom_sf(data = grats_yes, color = "#072b5f", linewidth = 1) +
  geom_sf(data = grats_no, color = "white", linewidth = 1) +
  coord_sf(
    xlim = c(bbox[c(1, 3)]) * 1.8,
    ylim = c(bbox[c(2, 4)]) * 1.4
  ) +
  labs(title = "Emblem of the Antarctic Treaty")
```

### Antarctica Flag Redesigned

In 2024, Graham Bartram presented a refreshed version of his flag that preserves
the white Antarctic outline but fills the blue field with many tiny, colorful
dots representing microplastics - even in remote regions. Because the design
relies on randomness, we approximate it using the following procedure:

1.  Sample random points across the Antarctic polygon.
2.  Build Voronoi polygons from those points, then apply a small negative buffer
    to create gaps.
3.  Randomly sample the resulting polygons to increase visual noise.
4.  Color polygons so larger areas remain white while smaller polygons use
    magenta/pink tones.

```{r 202510_redesign}
# Maximum chunk of Antarctica, the one that we fixed

ant_max_fixed

set.seed(2024)
# Sample, Voronoi and negative buffer
plastics <- st_sample(ant_max_fixed, 3000) |>
  st_union() |>
  st_voronoi(envelope = st_geometry(ant_max_fixed)) |>
  st_collection_extract() |>
  st_buffer(dist = -10000)


# Keep only those properly included in the outline

toinc <- st_contains_properly(ant_max_fixed, plastics, sparse = FALSE) |>
  as.vector()

# Select random chunks
plastic_end <- plastics[toinc, ] |>
  st_as_sf() |>
  slice_sample(prop = 0.75)

ggplot(plastic_end) +
  geom_sf(fill = "darkblue")


# Random coloring

plastic_end$area <- st_area(plastic_end) |> as.double()

plastic_end$fill <- sample(c("#ff00ec", "#9e00ec"), nrow(plastic_end), replace = TRUE)
plastic_end$fill <- ifelse(plastic_end$area > quantile(plastic_end$area, probs = 0.4),
  "white",
  plastic_end$fill
)

bbox2 <- st_bbox(plastic_end)
ggplot() +
  geom_sf(data = plastic_end, aes(fill = fill), color = NA) +
  scale_fill_identity() +
  theme(
    panel.background = element_rect(fill = "#009fdc"),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(title = "New redesign") +
  coord_sf(
    xlim = c(bbox2[c(1, 3)] * 1.8),
    ylim = c(bbox2[c(2, 4)]) * 1.4
  )
```
