---
layout: post
title: "Beautiful Maps with R (III): Patterns"
subtitle: "A solution for b/w and academic maps."
tags: [R,beautiful_maps, maps, sf]
date: 2019-06-02
share-img: https://dieghernan.github.io/assets/figs/20190602_imgpost-1.png
img-to-head: true
linktormd: true
output: github_document
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(global.par = TRUE)
rm(list = ls())


```

On this post I would introduce a couple of functions that may be useful for improving your maps. The goal is to produce different filling patterns that could be added over your shapefiles. 

At this point, I would like to suggest you (if you haven't done it yet) to install the `cartography`package ([vignette](https://cran.r-project.org/web/packages/cartography/vignettes/cartography.html)), as it presents a good bunch of interesting features, and fully compatible with the `sf` environment. In fact, I took some pieces of the base code in order to develop my own `legendPattern` function.

### Required R packages

```{r libs, warning=FALSE, message=FALSE, tidy='styler'}
library(sf)
library(rnaturalearth)
library(dplyr)
library(RColorBrewer)
```

### Grid approach

I already worked with the `st_make_grid` on a previous [post](https://dieghernan.github.io/201906_Beautiful1/), and this time I have applied the same approach. Some examples here on how the grid can be used to create patterns:

#### Dots

```{r 20190602_dotex, fig.align='center', message=FALSE, warning=FALSE, dpi=90, fig.width=200/90, fig.height=350/90,  paged.print=FALSE, tidy='styler', results='hide'}
DE  <- ne_countries(50,
                    type = "countries",
                    country = "Germany",
                    returnclass = "sf") %>%
  st_transform(3035)

grid = st_make_grid(DE,
                    what = "corners",
                    square = F)

# To avoid dots close to the edge
negbuff = st_buffer(DE, dist = -15 * 1000)
grid2 = grid[st_contains(negbuff, grid, sparse = F)]
par(mar = c(0, 0, 0, 0))
plot(st_geometry(DE))
plot(st_geometry(grid2), col = "red", add = T)
```

#### Grid

```{r 20190602_gridex, fig.align='center', message=FALSE, warning=FALSE, dpi=90,fig.width=200/90, fig.height=350/90, paged.print=FALSE, tidy='styler', results='hide'}
# Grid
grid = st_make_grid(DE,
                    what = "polygons",
                    square = T) %>%
  st_cast("LINESTRING") %>%
  st_intersection(DE)

# Clean and keep lines only
grid2 = grid[st_geometry_type(grid) %in% c("LINESTRING", "MULTILINESTRING")]

par(mar = c(0, 0, 0, 0))
plot(st_geometry(DE))
plot(st_geometry(grid2), col = "red", add = T)
```

Easy, right? Let's move to the next level!

#### Horizontal line

```{r 20190602_horex1, warning=FALSE, message=FALSE, tidy='styler', fig.width=200/90, fig.height=350/90, dpi=90, fig.align='center'}
# Horizontal
grid = st_make_grid(DE,
                    what = "polygons",
                    square = T)

par(mar = c(0, 0, 0, 0))
plot(grid)
plot(grid[55], add = T, col = "blue")
plot(
  st_point(st_coordinates(grid[55])[1, 1:2]),
  col = "red",
  add = T,
  pch = 16
)

plot(
  st_point(st_coordinates(grid[55])[2, 1:2]),
  col = "orange",
  add = T,
  pch = 16
)
plot(
  st_point(st_coordinates(grid[55])[3, 1:2]),
  col = "pink",
  add = T,
  pch = 16
)
plot(
  st_point(st_coordinates(grid[55])[4, 1:2]),
  col = "black",
  add = T,
  pch = 16
)
```

As it can be seen, we can extract specific points of each grid. Once that I identified them it is just a matter of connecting points using the `st_linestring` function.


```{r 20190602_horex2, warning=FALSE, message=FALSE, tidy='styler', fig.height=350/90,fig.width=200/90, dpi=90, fig.align='center'}

# Select horizontal only
grid_int = lapply(1:length(grid),
                  function(j)
                    st_linestring(st_coordinates(grid[j])[c(1, 2), 1:2])) %>%
  st_sfc(crs = st_crs(DE)) %>%
  st_intersection(DE)

# Clean and keep lines only
grid2 = grid_int[st_geometry_type(grid_int) %in% c("LINESTRING", "MULTILINESTRING")]

par(mar = c(0, 0, 0, 0))
plot(st_geometry(DE))
plot(st_geometry(grid2), col = "red", add = T)
```

### The `patternLayer` function

I put all that together on a function named `patternLayer` **PUT LINK HERE**. The main inputs are:

* `x`: `sf` object. It must be a `POLYGON` or a `MULTIPOLYGON`.
* `pattern`: fillings availables are `c("dot", "text", "diamond", "grid", "hexagon", "horizontal", "vertical", "zigzag", "left2right", "right2left")`.
* `density` of the grid. By default the function uses a grid with a minimum of 10 cells on the shortest dimension of the bounding box. Additionally it is possible to pass a `cellsize` value that would feed the `st_make_grid` underlying function.
* `txt` for the `text` pattern, that should be a character.
* Different graphical parameters can be also passed, as `add`, `cex`, `lty`, `pch`, etc. Basically three modes are allowed: points, lines and texts.

Let's see how it works.
```{r 20190602_showfun, warning=FALSE, message=FALSE, tidy='styler', fig.height=350/90,fig.width=500/90, dpi=90, fig.align='center'}

# Load function
source("patternfun.R")


par(mfrow = c(2, 5),
    mar = c(1, 1, 1, 1),
    cex = 0.5)
patternLayer(DE, "dot")
title("dot")
patternLayer(DE, "text", txt = "Y")
title("text")
patternLayer(DE, "diamond", density = 0.5)
title("diamond")
patternLayer(DE, "grid", lwd = 1.5)
title("grid")
patternLayer(DE, "hexagon", col = "blue")
title("hexagon")
patternLayer(DE, "horizontal", lty = 5)
title("horizontal")
patternLayer(DE, "vertical")
title("vertical")
patternLayer(DE, "left2right")
title("left2right")
patternLayer(DE, "right2left")
title("right2left")
patternLayer(DE, "zigzag")
title("zigzag")
```

Let's play a little bit more with some of the additional features of the function:

```{r 20190602_playing, warning=FALSE, message=FALSE, tidy='styler', fig.height=350/90,fig.width=500/90, dpi=90, fig.align='center'}


par(mar = c(1, 1, 1, 1), mfrow = c(2, 3))
plot(st_geometry(DE))
patternLayer(
  DE,
  "dot",
  pch = 10,
  density = 0.5,
  cex = 2,
  col = "darkblue",
  add = T
)
plot(st_geometry(DE))
patternLayer(
  DE,
  "dot",
  pch = 21,
  col = "red",
  bg = "green",
  cex = 1.25,
  add = T
)
plot(st_geometry(DE), col = "grey")
patternLayer(
  DE,
  "text",
  txt = "DE",
  density = 1.1,
  col = "white",
  add = T
)
plot(st_geometry(DE), col = "blue")
patternLayer(
  DE,
  "horizontal",
  lty = 3,
  cellsize = 150 * 1000,
  add = T
)
patternLayer(DE, "zigzag", lwd = 2, col = "red")
plot(st_geometry(DE), border =  "orange", lwd=2)
patternLayer(DE,
             "left2right",
             density = 2,
             col = "orange",
             add = T)
``` 