---
layout: post
title: Using CountryCodes database and sf package.
subtitle: Easy-to-use Database.
tags: [blog,R,sf,maps]
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())

```



This vignette is an example of use of the database provided in the Github project [Country Codes and International Organizations & Groups](https://github.com/dieghernan/Country-Codes-and-International-Organizations)
  by using the `sf` package in R.

##Required R packages

```{r libs, warning=FALSE, message=FALSE, tidy=TRUE}
library(sf)
library(jsonlite)
library(rnaturalearth)
library(dplyr)

```

## Reading the data
The first step consists on reading the database provided (in this example the `json` file) and extracting one international organization. In this example we will plot the [Commonwealth of Nations](https://en.wikipedia.org/wiki/Commonwealth_of_Nations).

```{r jsonandext, warning=FALSE, message=FALSE, tidy=TRUE}
df = fromJSON("https://raw.githubusercontent.com/dieghernan/Country-Codes-and-International-Organizations/master/outputs/Countrycodesfull.json")
#Identify Commonwealth acronym
orgsdb = read.csv("https://raw.githubusercontent.com/dieghernan/Country-Codes-and-International-Organizations/master/outputs/CountrycodesOrgs.csv") %>%
  distinct(org_id, org_name)
orgsdb[grep("Common", orgsdb$org_name), ]
```
 In our case, the value to search is **C**. It is provided also a function that extract the membership from the `json` database:
```{r funjson, warning=FALSE, message=FALSE, tidy=TRUE}
ISO_memcol = function(df,
                      orgtosearch 
                      ) {
  ind = match(orgtosearch, unlist(df[1, "org_id"]))
  or = lapply(1:nrow(df), function(x)
    unlist(df[x, "org_member"])[ind])
  or = data.frame(matrix(unlist(or)), stringsAsFactors = F)
  names(or) = orgtosearch
  df2 = as.data.frame(cbind(df, or, stringsAsFactors = F))
  return(df2)
}
df_org = ISO_memcol(df, "C")
```

Now `df_org` has a new column, named **C**, containing the membership status of each country.
```{r checkcol, warning=FALSE, message=FALSE, tidy=TRUE}
df_org %>% count(C)

df_org %>% filter(!is.na(C)) %>%
  select(ISO_3166_3,
         NAME.EN,
         C)  %>%   head()

```

## Replacing the data on a map.
In this example the `rnaturalearth` package is used for retrieving an `sf` object.  The code below replaces the `data.frame` part of the `sf`object.
and replacing the `dataframe`for the dedicated database.

```{r mapandrep, warning=FALSE, message=FALSE, tidy=TRUE}
testmap = ne_countries(50,
                       "countries",
                       returnclass = "sf") %>% select(ISO_3166_3 = adm0_a3) %>%
  full_join(df_org)
```

##Plotting map: Wikipedia style
Now we would try to plot a map ressembling the one presented in the [Wikipedia page](https://en.wikipedia.org/wiki/Commonwealth_of_Nations) for the Commonwealth.
```{r download, include=FALSE}
download.file("https://upload.wikimedia.org/wikipedia/commons/e/e2/Member_states_of_the_Commonwealth_of_Nations.svg",
              "wikiCommonW.svg")
```
<img src="wikiCommonW.svg?sanitize=true">

The map we will generate is presented under a Robinson projection and the color palette will be based in the [Wikipedia convention for Orthographic Maps](https://en.wikipedia.org/wiki/Wikipedia:WikiProject_Maps/Conventions/Orthographic_maps), since it is the one used in the example.

```{r mapfin, warning=FALSE, message=FALSE, tidy=TRUE}
#Projecting the map
testmap_rob = st_transform(testmap, "+proj=robin")

#Bounding box
bbox = st_linestring(rbind(c(-180, 90),
                           c(180, 90),
                           c(180, -90),
                           c(-180, -90),
                           c(-180, 90))) %>%
  st_segmentize(5) %>%
  st_cast("POLYGON") %>%
  st_sfc(crs = 4326) %>%
  st_transform(crs = "+proj=robin")

#Plotting
par(mar = c(0, 0, 0, 0))
plot.new()
plot(bbox,
     col = "#FFFFFF",
     border = "#AAAAAA",
     lwd = 1.5)
plot(
  st_geometry(testmap_rob),
  col = "#B9B9B9",
  border = "#FFFFFF",
  lwd = 0.1,
  add = T
)

plot(
  st_geometry(testmap_rob %>%
                filter(!is.na(C))),
  col = "#346733",
  border = "#FFFFFF",
  lwd = 0.1,
  add = T
)

```
