---
layout: page
title: "Pretty labs"
subtitle: "Vignette of getFormatNums"
tags: [R, sf, cartography, vignette]
date: 2020-03-10
permalink: /prettylabs/
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false 
    toc_depth: 1
vignette: >
  %\VignetteIndexEntry{Pretty labs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

source code: [here](https://github.com/dieghernan/dieghernan.github.io/blob/master/blog/prettylabs.Rmd)

  
```{r echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
rm(list = ls())
knitr::knit_hooks$set(margin = function(before, options, envir){
  if (before){
    par(mar=c(0.1,0.1,1.3,0.1))
  } 
})
```


# Installation
```{r install, message=FALSE, warning=FALSE}
library(devtools)

devtools::install_github("dieghernan/cartography", ref="prettylabs")

library(cartography)



```

# man
## `getFormatNums`: Pretty format for labels

### Description


 Convert numeric arrays into formatted strings.


### Usage

```r
getFormatNums(
  v,
  thousands = "",
  decimals = getOption("OutDec"),
  values.rnd = 0,
  prefix = "",
  suffix = "",
  align = "left",
  leadzero = TRUE,
  replace.zero = NULL
)
```


### Arguments

Argument      |Description
------------- |----------------
```v```     |     a vector of numeric values.
```thousands```     |     thousands separator.
```decimals```     |     decimals separator.
```values.rnd```     |     desired number of digits after the decimal separator.
```prefix, suffix```     |     strings to be pasted before of after the number, see Details.
```align```     |     alignment of the final string, possible values are "left" or "right".
```leadzero```     |     logical, convert leading zeros to `" ."` for values in (-1,1).
```replace.zero```     |     replace zero values for this parameter. See Details.

### Details


 If `suffix` contains `"%"` the values are converted to percentages.
 
 `NA` values of `v` are converted to `"NA"` . 
 
 If `replace.zero` is not `NULL` , exact zeroes would be replaced for this value.


### Value


 A character vector containing the formatted values.


### Author


 dieghernan, [https://github.com/dieghernan/](https://github.com/dieghernan/) 


### Examples

```{r examples, dev='svg', fig.height=7, fig.width=7, margin=TRUE}
library(sf)
mtq <-
  st_read(system.file("gpkg/mtq.gpkg", package = "cartography"))
# Population density
mtq$POPDENS <- 1e6 * mtq$POP / st_area(x = mtq)
brks <- getBreaks(mtq$POPDENS, 6)
cols <- carto.pal("red.pal", 6)
labs <- getFormatNums(brks,
                      thousands = ",",
                      suffix = " in/km2",
                      align = "right")

choroLayer(mtq,
           var = "POPDENS",
           col = cols,
           legend.pos = "n")
legendChoro(
  breaks = labs,
  col = cols,
  title.txt = "Population Density",
  pos = "topright"
)

# Percentage of population
mtq$PORCPOP <- mtq$POP / sum(mtq$POP)

brks <- seq(-1, 1, .2)
brks <- c(brks,NA)
cols = carto.pal("sand.pal", 20)
labs <- getFormatNums(
  brks,
  values.rnd = 2,
  leadzero = FALSE,
  replace.zero = "zero"
)

choroLayer(mtq,
           var = "PORCPOP",
           col = cols,
           legend.pos = "n")

legendChoro(
  breaks = labs,
  col = cols,
  title.txt = "Population Distr.",
  pos = "topright",
  nodata = FALSE
)

# Area percent
mtq$AREA <- (st_area(mtq) / sum(st_area(mtq)))

brks <- getBreaks(mtq$AREA, method = "q6")
cols = carto.pal("orange.pal", 6)
labs <-
  getFormatNums(brks,
                suffix = "%",
                values.rnd = 1,
                prefix = "<")
choroLayer(mtq,
           var = "AREA",
           legend.pos = "n",
           col = cols)

legendChoro(
  breaks = labs,
  col = cols,
  title.txt = "% Area",
  pos = "left",
  nodata = FALSE
)

```

 
# `discLayer` and `legendGradLines`
 
```{r discl, dev='svg', fig.height=7, fig.width=7, margin=TRUE}

library(sf)
mtq <-
  st_read(system.file("gpkg/mtq.gpkg", package = "cartography"))
# Get borders
mtq.borders <- getBorders(x = mtq)
# Median Income
choroLayer(
  x = mtq,
  var = "MED",
  border = "grey",
  lwd = 0.5,
  method = 'equal',
  nclass = 6,
  legend.pos = "topleft",
  legend.title.txt = "Median Income\n(in euros)"
)
# Discontinuities
discLayer(
  x = mtq.borders,
  df = mtq,
  var = "MED",
  col = "red4",
  nclass = 3,
  method = "equal",
  threshold = 0.4,
  sizemin = 0.5,
  sizemax = 10,
  type = "abs",
  legend.values.rnd = 0,
  legend.title.txt = "Discontinuities\n(absolute difference)",
  legend.pos = "bottomleft",
  add = TRUE,
  prefix = "$ ",
  suffix = " this is a long suffix"
)

```

# `legendCirclesSymbols` (and partially `propSymbolsLayer`)

```{r legC, dev='svg', fig.height=7, fig.width=7, margin=TRUE}

library(sf)
mtq <- st_read(system.file("gpkg/mtq.gpkg", package="cartography"))
plot(st_geometry(mtq))
box()

propSymbolsLayer(x = mtq, var = "POP",
                 inches = 0.2, 
                 thousands = " ",
                 decimals =",")


legendCirclesSymbols(pos = "right", inches = 0.2,
                     var = -c(min(mtq$POP), max(mtq$POP))/10000,
                     values.rnd= 4,
                     suffix = "%")
legendCirclesSymbols(pos = "left",
                     title.txt="ES",
                     cex=2,
                     var = c(min(mtq$POP), max(mtq$POP)),
                     inches = 0.2, style = "e",
                     prefix = "pop     is ")

```
 
 

